---
title: "lab2"
output: html_document
date: "2022-11-25"
---

```{r}
library(ltm)
```
1. Load and display the dataframe Abortion (library ltm).
```{r}
data("Abortion")
dim(Abortion)
head(Abortion)

```

2. Use the function descript to summarize the principal characteristics of the data set dataset. In particular evaluate:
- which one among the 4 items shows the lowest degree of agreement among the sampled individuals; 
- what is the distribution of the total score in the sample; 
- if there are problems in some pairs of items ;
```{r}
?descript # performs descriptive analysis on this type of data.

dsc<-descript(Abortion)
dsc$perc
```
percentage of respondent for each item: 
56% of people answer 0 to item 1 and so the remaing 44% answer 0.
The logit of the third columns is the log odds.
So the item with the lowest degree of agreement is the first with 43%

```{r}
dsc$item
```
103 respondents respond 0 to all questions.
141 , one to all questions.
33, one to only one question.
...

```{r}
dsc$pw.ass
```
All pvlaues are very small and this means that all items are associated among each other in pairs.
Therefore there are not problems.
if some pvlaue will be >0.05 it means that it will be significant so the correspondent items will be not associated.


2. Use the function ltm to fit the 2 IRT model to the data. Comment the results. In particular describe :
2.1 Which item shows the highest positive attitude towards the abortion by the interviewed individuals
```{r}
m1<-ltm(Abortion~z1)
m1.rip<-ltm(Abortion~z1, IRT.param = F)
summary(m1)
summary(m1.rip)
m1.rip$coefficients[,1]
```

2.2 Which item more discriminate among individuals in favour and against the abortion
```{r}
summary(m1.rip)
alpha<-m1.rip$coeff[,2]
```
Item 3

2.3 Which item less discriminate among individuals in favour and against the abortion
```{r}
alpha<-m1.rip$coeff[,2]
alpha
```
Item 4


2.4 The interpretation of the latent variable
```{r}
summary(m1.rip)
```
(Lavaan does not fit more than 2 latent variables.)
The first 4 parameters are the difficulty parameters and they are written as intercepts.
  the first one is the most difficult (the smaller the more difficult).
  
  
The last 4 parameters are the discriminant parameters and the one that discriminate more are the third.
```{r}
summary(m1)
plot(m1.rip)
plot(m1)
```
In the GLLMV parameters can be interpreted as loadings(in IRT framework not).
Bi0= - ai0 / aij
So since the sign is the opposite, the higher the difficulty parameter the higher the difficulty to have a correct response on the item.
with ITR parameter = TRUE we deal with the 2parametric logistic model so difficulty parameter have a different interpretation (opposite). Here the discrimination parameter are equal to the 2 parametric logistic model.


2.5 Verify the relation between the parameter estimates in the two parameterizations.
```{r}
summary(m1)$coef[1,1]  #[1] 0.1696615
summary(m1.rip)$coef[1,1]  #[1] -0.7555329

-summary(m1)$coef[1,1]*summary(m1)$coef[5,1]   #[1] -0.7555329
-summary(m1.rip)$coef[1,1]/summary(m1)$coef[5,1]  #[1] 0.1696615
```


4.3 Compute the standardized discriminant parameters and the probability of positive response of the median individual
```{r}
alpha<-m1.rip$coeff[,2]
stalpha<-alpha/sqrt(1+alpha^2)

coef(m1.rip, prob=TRUE)
coef(m1.rip, prob=TRUE, order=TRUE)
#coef(m1, prob=TRUE, order=TRUE)
```
By definition of the parametrisation we need the standardized alpha.ij
the item with the highest probability of positive response of the median individual is the thir, coherent as before.


6.Represent the characteristic curves of the four items for the estimated model. Comment the results.
```{r}
plot(m1,legend=TRUE, cx= "bottomright", 
     xlab="attitude toward abortion", lwd=3, cex.main=1.5, 
     cex.lab=1.3, cex=1.1)

```

Since we have a logit model we obtain the logistic curve and it is useful when we have a lot of items.


7. Based on the obtained results evaluate if the model has a good fit to the data.
```{r}
E<-fitted(m1)[,5]
O<-m1$patterns$obs

1010
1001

cbind(m1$patterns$X, O, E)

chisq<-sum((O-E)^2/E)  #in the application i have to consider only the observed patterns so it is not 2^p with p=16 but with p=14.

#2^p=14 observed pattern
14-2*4-1
df<-14-4-4-1
pvalue<-pchisq(chisq, df, lower.tail = F)
1-pchisq(chisq,df)

LR<-2*sum(O*log(O/E)) #likelihood ratio test
pvalueLR<-1-pchisq(LR, df)
```
The object fitted will contain the expected frequencies 
We obtain all the possible response pattern that we can have but two of them that are not observed (14 instead of 16).
Some expected frequencies are very low<1, this means that data will be sparse and the statistic tests reject the hypothesis even if the model is adequate.

8. Use the function margins to compute the values of the two-way and three-way chi-square residuals and evaluate if the model has a local goodness of fit to the data.
```{r}

margins(m1)
margins(m1, type="three-way", nprint=2)
margins(m1, type="three-way", nprint=3)
```
The last column is the Rstatistics.
(m1)There are not bivariate residuals greater than 4. so they are good.
(m1, 2) We decide to remove the item with R value >4.
Since we have find out that the data are spare we look to the residuals calculated from marginal frequencies.

9. Use the function factor.scores to estimate the values of the latent variable for each individual.
```{r}
?factor.scores
fs<-factor.scores(m1, method = "EAP")
comp<-factor.scores(m1, method = "Component")
total.score<-apply(fs$score.dat[,1:4], 1, sum) #sum
resp.pattern<-fs$score.dat[,1:4]
total.score<-apply(resp.pattern,1,sum)
cp<-comp$score.dat[,7]
tab<-cbind(fs$score.dat,cp,total.score)
tab
```
The discriminant parameters are all positive. The first score is zero because all values of x are 0. (comp)
Total score shows the number of positive responses of individuals for each response pattern.
Here latent variable is the attitude toward abortion. (tab)
This method gives the same ranking of the individual, so we can order this table according to the total score and see what happen to the other methods score.

```{r}

round(tab,3)
round(tab[order(total.score),],)
```

z1 and cp have exactly the same order because the component is a sufficient statistic for the latent variables.
The total score method gives the same ranking of the other methods.



10. Estimate the abilities of the response pattern that do not occur in the dataset.
```{r}
fitted(m1,resp.pattern=rbind(c(1,0,0,1), c(1,0,1,0)))
factor.scores(m1,resp.pattern=rbind(c(1,0,0,1), c(1,0,1,0)))
```
Response pattern that we have not observed.
value of the expected frequencies for these two patterns. (ther are very low expected frequencies)
