---
title: "lab3 ex2"
output: html_document
date: "2022-12-30"
---

1. Load the dataset MObility from library lmt
```{r}
library(ltm)
data(Mobility)
head(Mobility)
dim(Mobility)
?Mobility
```

2. Recode the data set properly in order to use the function poLCA.
```{r}
names(Mobility)<-c("Item1","Item2","Item3","Item4", "Item5","Item6","Item7","Item8")
Mobility[Mobility==1]<-2
Mobility[Mobility==0]<-1
```


3. Use the function poLCA to fit the models with 2, 3 e 4 latent classes to the data.
```{r}
library(poLCA)
formula<-cbind(Item1,Item2,Item3,Item4,Item5,Item6,Item7,Item8)~1
m.2<-poLCA(formula,Mobility,nclass=2,nrep=10,verbose=FALSE)
m.3<-poLCA(formula,Mobility,nclass=3,nrep=10,verbose=FALSE)
m.4<-poLCA(formula,Mobility,nclass=4,nrep=10,verbose=FALSE)
```


4. Compare the estimates of the expected frequencies for the response pattern obtained with the three models.
```{r}
freq.estim<-data.frame(m.2$predcell[1:10],m.3$predcell[10],m.4$predcell[10])
freq.estim
```
The table results that the expected frequencies obtained by the three methods are very different from each other and also with respect to the observed frequencies. 
Moreover there are some expected frequencies that are <1 and this shows that we are dealing with sparse data.


5. Compare the three models considering the following measures: log-likelihood, number of free parameters, likelihood ratio test, Pearson chi-square test (and associated p-value), AIC, BIC.
```{r}
K<-c("2", "3", "4")
llik<-c(m.2$llik,m.3$llik,m.4$llik)
npar<-c(m.2$npar,m.3$npar,m.4$npar)
Gsq<-round(c(m.2$Gsq,m.3$Gsq,m.4$Gsq),3)
Chisq<-round(c(m.2$Chisq,m.3$Chisq,m.4$Chisq),3)
df<-c(m.2$resid.df,m.3$resid.df,m.4$resid.df )
pvalue<-round(1-pchisq(Chisq,df ),4)

AIC<-round(c(m.2$aic,m.3$aic,m.4$aic),3)
BIC<-round(c(m.2$bic,m.3$bic,m.4$bic),3)
summary<-data.frame(K,llik,npar,Gsq,Chisq,df,pvalue,AIC,BIC)
```

6. Which model shows the best fit to the observed data?
According to the AIC and to the BIC the 4classes model is the one that provides the best fit to the data.


7. Display the parameter estimates of the four-class model and compare this solution with those obtained with your colleagues. Which problem arises?
```{r}
round(m.4$P,4)
lapply(m.4$probs,round,4)
```
The class are not order because of the seed set in the computer.


8. Use the function poLCA.reorder to solve the indeterminacy of the ranking of the
latent classes.
```{r}
probs.start.m4<-m.4$probs.start
new.probs.start.m4<-poLCA.reorder(probs.start.m4,order(m.4$P))
m.4.ord<-poLCA(formula,Mobility,nclass=4,probs.start=new.probs.start.m4 ,verbose=FALSE)
round(m.4.ord$P,4)
lapply(m.4.ord$probs,round,2)
```


9. Interpret the four latent classes.
1class: is about women that have a very high mobility, yes to all the quesitons.
2class: women that have a medium level of mobility, in and outside the town but that do not love doing collective actitivies.
3class:the class in which the women never did anything, no to all the questions.
4class: women with a medium mobility that went inside and outside the town but without activities for getting in touch with other people.


10. Display the posterior probability estimates of the response pattern (1,1,1,1,1,1,1,1) and (2,2,2,2,2,2,2,2) for the four latent classes. In which classes the samples of women corresponding to these response patterns are allocated?
```{r}
a<-as.numeric(row.names(Mobility[which(Mobility$Item1==1&Mobility$Item2==1&
Mobility$Item3==1&Mobility$Item4==1&Mobility$Item5==1&Mobility$Item6==1&
Mobility$Item7==1&Mobility$Item8==1),]))
post.1<-m.4.ord$posterior[a[1],]
round(post.1,2)
b<-as.numeric(row.names(Mobility[which(Mobility$Item1==2&Mobility$Item2==2&
Mobility$Item3==2&Mobility$Item4==2&Mobility$Item5==2&Mobility$Item6==2&
Mobility$Item7==2&Mobility$Item8==2),]))
post.2<-m.4.ord$posterior[b[1],]
round(post.2,2)

```
```{r}
round(m.4.ord$posterior[185,],5)
```


11. Display the latent class in which the samples of women are allocated according to highest posterior probability.
```{r}
m.4.ord$predclass
table(m.4.ord$predclass)

```


12. Select the samples of women allocated to the class with less mobility and evaluate the corresponding response patterns.
```{r}

cl3<-Mobility[m.4.ord$predclass==3,]
cl3
#unique(cl3)
```

13. Compute the predicted cell probability from the latent class model of the response pattern (2,1,2,1,2,1,2,1), that is not observed in the dataset, and allocate it according to the highest posterior probability.
```{r}
poLCA.predcell(m.4.ord,c(2,1,2,1,2,1,2,1))
round(poLCA.posterior(m.4.ord,y=c(2,1,2,1,2,1,2,1)),3)
```
According to its posterior probability, the pattern will be allocated in the second class.
